Drought, vegetation productivity, and land cover change in Chile

### 2024-03-12 article/paper_drought_Chile2023.qmd
update paper qm intro

title: Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS

@@

  # - name: Bob Security
  #   email: bob@example.com
  #       - id: another-u
  #         name: Another University
  #         department: Department Name
  #         address: Street Address
  #         city: City
  #         state: State
  #         postal-code: Postal Code

@@

 Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a “Mega Drought". Our objective is to examine the effects of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and vegetation productivity. For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande," “Norte Chico," "Centro," "Sur," and "Austral." The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used. From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1. We estimated atmospheric evaporative demand (AED) using the Hargreaves-Samani equation with the ERA5L temperature. We used the Standardized Precipitation Index (SPI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized anomaly of cumulative NDVI (zcNDVI) as drought indicators. These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months. We analyze the trend for LULCC, vegetation productivity, and drought indices. Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity. Our results showed that LULCC were highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34%, respectively. The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply. Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." Showing to be the zones most impacted by climatic conditions, the years 2020 and 2022 suffered the most extreme drought. On the other hand, forests seem to be the most resistant to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.

@@

      cite-style: number

@@

The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013]. There is a high degree of confidence that rising temperatures will increase the land area where droughts will occur more frequently and with greater severity [@IPCCCH112021]. Furthermore, drought increases tree mortality and triggers changes in land cover and, consequently, land use, thus impacting ecosystems [@Crausbay2017]. Nevertheless, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

The primary cause of drought is precipitation, and temperature makes it worse [@Luo2017]. Drought impacts soil moisture, hydrological regimes, and vegetation productivity. Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought. Even though it has been argued that those definitions do not fully address the ecological dimensions of drought, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems.". Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. On the other hand, the AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.

Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted vegetation development [@Zambrano2023] and the hydrological system [@Boisier2018]. Current drought conditions have affected crop productivity [@Zambrano2016; @Zambrano2018], forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], land cover change [@Fuentes2021], water supply in watersheds [@AlvarezGarreton2021], and have had economic impacts [@Fernandez2023]. In 2019–2020, the drought severity reached an extreme condition in Central Chile (30–34°S) not seen for at least 40 years, and the evidence indicates that the impact is transversal to the land cover classes of forest, grassland, and cropland [@Zambrano2023]. The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.

For the spatiotemporal assessment of drought impact (i.e., by water supply and demand) on land cover changes, we need climatic realiable variables such as precipitation, temperature, soil moisture, land cover, and vegetation status. For developing countries like Chile, the weather networks present several disadvantages, such as gaps, a short history, and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension. ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil moisture and evapotranspiration. On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics. The Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get vegetation drought indices (VDI), which are often used as proxies for productivity [@Paruelo2016; @Schucknecht2017]. Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021]. To analyze these changes, multiple LULC products exist [@Grekousis2015]. One of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS. The variation in water supply and demand is finally reflected in the total water storage (TWS). The Gravity Recovery and Climate Experiment (GRACE), which allows analyzing changes in water availability at coarse resolution, can retrieve the TWS [@Ahmed2014; @Ma2017]. We can find drought indices of supply (i.e., precipitation) and demand (i.e., temperature) using climatic reanalysis (ERA5L) and vegetation data (MODIS). This lets us figure out how drought changes LULC. Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.

To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012) recommends the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term. Following the same approach, @Vicente-Serrano2010 incorporates into the SPI the effect of temperature through the use of potential evapotranspiration, thus proposing the SPEI (Standardized Precipitation Evapotranspiration Index). Similarly, to evaluate solely the evaporative demand driven by temperature, @Hobbins2016 and @McEvoy2016 came up with the Evaporative Demand Drought Index (EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI; @Zambrano2018 proposed the zcNDVI, a standardized anomaly of the cumulative Normalized Difference Vegetation Index (NDVI), which could be acumulated over the growing season or any period (e.g., months), resulting in a multiscalar drought index. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a normalized index [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021] which is a normalized index using vegetation, land surface temperature, and a vegetation condition index (VCI, [@Kogan1995]). From TWS, we can estimate the standardized terrestrial water storage index (STI) [@Cui2021], a standardized anomaly that follows the methodology of the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for water supply, demand, and storage, which can help to make a comprehensive assessment of drought.

In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile. Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2022 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop). The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009]. To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the land cover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc). "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb). There is a significant amount (50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), then grassland (16%), forest (8%), and croplands (5%), in "Centro." The south part of "Sur" and the north part of "Austral" are dominated by an oceanic climate (Cfb). Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

```{r fig-studyArea}
#| echo: false
#| fig-pos: "!ht"
#| fig-env: 'figure*'
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral". (b) Topography reference map. (c) Land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022'
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/map_study_con_landcover.png")
```

@@

For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023]. For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019]. For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023]. For managing vectorial data, we used `{sf}` [@Pebesma2018]. For the calculation of AED, we used `{SPEI}` [@Bergueria2023].

@@

To derive a proxy of vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.

@@

We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations. For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023. We used 277 stations located throughout Chile. For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops. We used daily data for the year 2022 and the months outside the growing season, May to September, to avoid the effect of irrigation on soil moisture, which is hardly captured by ERA5L.

@@

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993]. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses soil moisture at 1m depth. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed using a locally-weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time-series with the ten classes for 2001 and 2023.

@@

i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
iii) calculate a confusion matrix with the classes extracted in the 1000 poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

@@

The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type. In this study, we want to analyze the variation in vegetation productivity in different land cover types and how it is affected by water demand, water supply, and soil moisture. In order to avoid changes due to a change in the land cover type, that will wrongly impact NDVI. We will develop a persistence mask for land cover for 2001–2023. Thereby, we reduce an important source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no land cover change for the macroclasses.

@@

We analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale. Then, we will summarize for each land cover class and macrozone. Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses having vehetation: forest, cropland, grassland, savanna, and shrubland.

To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI. We followed a similar approach to that used by @Meroni2016 when using the SPI for meteorological drought against the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation) as a proxy for vegetation productivity. We made a pixel-to-pixel linear correlation analysis per index. First, we calculate the Pearson coefficient of correlation for the six time scales and let the time scale that reaches the maximum correlation be significant (p < 0.05). Then, we extracted the Pearson correlation value corresponding to the time scales that reached the maximum value. Thus, we derived two raster maps per index, the first with the time scales and the second with the correlation value.

<!-- ##	Total water storage and drought indices -->

<!-- Terrestrial Water Storage (TWS) is defined as the total amount of water stored on land that includes any natural or artificial water bodies, such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and biomass water [@Humphrey2023; @Deng2023]. TWS changes evidence the effects of multiple water fluxes on the hydrological cycle [@Deng2023]. These are reflected in the temporal variation of observations of the Earth's gravitational field [@Abolafia2021; @Sabzehee2023] and, in recent decades, the twin Gravity Recovery and Climate Experiment (GRACE) satellites and their Follow-On mission (GRACE-FO) have provided valuable results on globally distributed TWS anomalies [@Tapley2019; @Ferreira2023]. The GRACE mission was launched in March 2002 and was operational until October 2017 [@Ramjeawon2022]. Then its GRACE Follow-On successor was launched in May 2018 [@Landerer2020; @Yin2022]. The information provided by these satellites is used to construct monthly maps of the Earth's average gravity field, providing details of the movement of water masses or water mass anomaly estimates relative to the long-term average gravity field [@Humphrey2023; @Wahr2004]. -->

<!-- In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022]. This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@NASA2019; @Wiese2016]. Although the mascon solutions greatly reduced leakage errors, a gain factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022]. The water storage and height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and their temporal resolution monthly is from April 2002 to May 2023. -->

@@

\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-land coverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

>>>>

title: Drought, vegetation productivity, and land cover change in Chile

@@

  - name: Anton Vrieling
    email: a.vrieling@utwente.nl
        - id: itc
          name: University of Twente
          department: Faculty of Geo-Information Science and Earth Observation (ITC)
          city:  Enschede, The Netherlands
  - name: Francisco Fernández
    email: a.vrieling@utwente.nl
        - id: uss
          name: Universidad San Sebastian
          department: 
          city:  
  - name: Francisco Meza
  - name: Alejandro Venegas-González
  - name: Iongel Duran-Llacer
  - name: Nicolas Raab
  - name: Dylan Craven        

@@

  Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a megadrought. Our objective is to assess the impact of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and their impact on vegetation productivity. The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used. From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1. As drought indices, we compute the standardized anomaly of cumulative NDVI (zcNDVI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), and the Standardized Soil Moisture Index (SSI). These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months. We analyze the trend for LULCC, vegetation productivity, and drought indices. Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity. Our results showed that LULCC was highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34% of change in the surface type, respectively. The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply. Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." On the other hand, forests seem to be the most resistant to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits lasting longer than a year. Our results can help develop a robust vegetation productivity forecasting model for land cover classes in Chile.

@@

@@

Drought is often classified as meteorological when there is a decrease in precipitation below the mean average of several years (more than 30 years), hydrological when these anomalies last for long periods (months to years) and affect water systems, and agricultural when the deficit impacts plant health anomalies and leads to decreased productivity [@Wilhite1985]. However, it is important to note that drought is also influenced by human activities, which were not considered in the definitions. Thus, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans’ decisions and activities that drives the anthropogenic drought. Simultaneously, drought leads to heightened tree mortality and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017]. Even though many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. Then, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems." In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on water security. [@Bakker2012]

Human-induced greenhouse gas emissions have increased the frequency and/or intensity of drought as a result of global warming, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, suggesting that drought has not exhibited a significant trend over the past forty years. [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the meteorological drought trend on a global scale, finding that only in a few regions has there been an increase in the severity of drought. Moreover, they attribute the increase in droughts over the past forty years solely to an increase in atmospheric evaporative demand (AED), which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that "the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change". Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years. Further, the @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments are needed that relate meteorological and soil moisture variables to their effects on vegetation.

From 1960 to 2019, land use change has impacted around one-third of the Earth's surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021; @Song2018] and regionally [@Chamling2020; @Homer2020; @Yang2021]. Some others seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on the interaction between drought and land cover change [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 conducted a worldwide investigation utilizing net primary production to examine the spatial and temporal variations in vegetation productivity at global level. The study aimed to assess the influence of drought by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline in vegetation productivity, while water availability accounts for 55% of the variation. @Chen2022 studied the trend of vegetation greenness and productivity and its relation to meteorological drought (SPEI of twelve months in December) and soil moisture at the global level. The results showed lower correlations (<0.2) for both variables. @Akinyemi2021 evaluates drought trends and land cover change using vegetation indices in Botwsana in a semi-arid climate. These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) over a single time period of 12 months. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own. Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological droughts? ii) How do they behave in humid and arid climatic zones regarding drought? And iii) What is the role of soil moisture? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

Chile's diverse climatic and ecosystem types (@Beck2023;@Luebert2022) make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. Central Chile faced a persistent precipitation deficit between 2010 and 2022, defined as a megadrought [@Garreaud2017], which has impacted the Chilean ecosystem. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016]. We found one study regarding land cover and drought in Chile. The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude. @Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results. For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI. The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Hence, it is imperative to assess temporal scales that take into account the cumnulative effect within some years. There is little knowledge about the relationship between drought and ecosystem in Chile; thus, it is important to understand in more detail how meteorological and soil moisture droughts influence ecosystem dynamics to inform adaptation options.

A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture. Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially. In order to do this, we use reanalysis data from ERA5-Land [@MunozSabater2021] to create drought indices that consider AED, precipitation, and soil moisture over a range of time periods, from the short to the long term. Also, we use vegetation spectral information and annual land cover change from the Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity. Here, we analyze the multi-dimensional impacts of drought across ecosystems in continental Chile. More specifically, we aim to assess: i) temporal changes in land-use cover and the direction and magnitude of their relationships with drought indices for water demand and supply, soil moisture, and vegetation productivity; ii) short- to long-term temporal trends in multi-scalar drought indices; and iii) the relationship between vegetation productivity and drought indices for water demand and supply and soil moisture across Chilean ecosystems.

@@

Continental Chile has a diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity (@Luebert2022) (@fig-studyArea c). The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009]. In order to characterize the climate and ecosystem of Chile, we utilize the Köppen-Geiger classification system developed by @Beck2023 and the land cover data derived from the MODIS product for the period of 2001–2022, based on the International Geosphere-Biosphere Programme (IGBP) classification scheme proposed by @Friedl2019. "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in "Austral", it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

```{r fig-studyArea}
#| echo: false
#| fig-pos: "!ht"
#| fig-env: 'figure*'
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral". (b) Topography reference map. (c) land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022'
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/map_study_con_landcover.png")
```

@@

For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023]. For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019]. For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023]. For managing vectorial data, we used `{sf}` [@Pebesma2018]. For the calculation of AED, we used `{SPEI}` [@Bergueria2023]. For mapping, we use {tmap} [@Tennekes2018]. For data analysis, the suite {tidyverse} [@Wickham2019] was used.

@@

To derive a proxy for vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.

@@

We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations. For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023. We used 277 stations located throughout Chile. For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops. The sensors are installed at 30, 60, and 90m and are the model Teros 12 from MeterGroup. To avoid the effect of irrigation on soil moisture, which ERA5L hardly captures, we used daily data for the year 2022 and the months outside the growing season, May to September.

@@

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. We used the SPI [@Mckee1993], a drought monitoring index that the World Meteorological Organization (WMO) recommends. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses soil moisture at 1m depth. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed using a locally weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.

@@

i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
iii) calculate a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

@@

Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI. In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it. In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we will develop a persistence mask for land cover for 2001–2023. Thereby, we reduce an important source of variation on a regional scale.

Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remain constant for at least 80% of the years (2001–2022). This enables us to identify regions where the land cover of the macroclasses has remained constant.

@@

We analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale. Then, we will summarize for each land cover class and macrozone. Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses of vegetation: forest, cropland, grassland, savanna, and shrubland.

An analysis is conducted on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI. The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity. We implemented a methodology similar to that of @Meroni2016 when comparing the SPI for meteorological drought to the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation), which served as an indicator for vegetation productivity. A pixel-to-pixel linear correlation analysis was performed for each index. To begin, the Pearson coefficient of correlation is computed for each of the six time scales. A significant time scale is identified as the one that attains the highest correlation (p < 0.05). Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted. As a result, for each index, we generated two raster maps: one containing the time scales and the other the correlation value.

@@

\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-landcoverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}
